---
title: 'Exploratory analysis of Atlantic hurricanes'
author: "Rohit Tahade"
date: "June 2021"
output:
  html_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r}
#calling required libraries
library(tidyverse)
library(maps)
#Loading hurricane data
# Load the data file using a relative path
load("./processed_data/Hurdat2_tidy.RData")


```

## Question 1 : What's in this dataset?

#### a. Load the dataframe. Pick one famous hurricane that you know the name and make a dataframe similar to hurrs but containing only the data for your chosen hurricane. (Note that the people who name hurricanes have reused names sometimes.)

*Chose hurricane MARIA for analysis & filtering for id AL14 for study*

```{r}
hurrs %>% filter(name=="MARIA") %>% filter(str_detect(id,"AL14")) -> maria
```








#### b.	Write a function called Hurdat2_summary that takes the hurrs dataframe as an input and returns a named list containing
•	the total number of storms
•	the range of years covered
•	a vector containing the number of track points for each storm


```{r}
Hurdat2_summary <- function(df){
  #count the total number of storms
  total_storms<- summarise(df,count=n())
  
  #range of years covered
  range<-summarise(df,from=min(year),To=max(year))
  
  #a vector containing the number of track points for each storm
  track_points<- group_by(df,name) %>% summarise(total_track_pt=n()) 
  
  
  track_points <- setNames(as.numeric(track_points$total_track_pt),as.character(track_points$name))
  list("Total no. of Storms"=total_storms,"Range of years covered"=range,"Track Points of each storm"=track_points)
}

```

\*Test your function by running it for three cases:
•	the full hurrs dataframe
•	a version that has been filtered to contain only storms from a single year
•	the one-hurricane dataframe you made in (a)


-   Test 1

    ```{r}
    Hurdat2_summary(hurrs)
    ```

-   Test 2

    ```{r}
    Hurdat2_summary(filter(hurrs,year==2005))
    ```

-   Test 3

    ```{r}
    Hurdat2_summary(maria)
    ```








#### c.	Make a histogram of the number of track points, using the output of your function for the hurrs case.


```{r}
 #converted list to a data frame 
histplot<-as.data.frame(Hurdat2_summary(hurrs)) 
# Taking rownames into name column
histplot<-rownames_to_column(histplot,"name") 
# filtering out unnamed storms and plotting histogram
histplot %>% filter(name!="UNNAMED") %>%     
             ggplot() +                      
             geom_histogram(aes(x=Track.Points.of.each.storm),alpha=0.5,color="white", fill="red") + 
             labs(title = "No. of Track points of storm all storms", x= "Track points")
```







#### d.	What year did they begin naming hurricanes and other storms?

```{r}
#Arranging as per year
hurrs %>% arrange(year) %>% 
#filtering out unnamed storms 
filter(name!="UNNAMED") %>%
#selecting first named entry and converting to numeric vector
head(1) %>% select(year) %>% as.numeric() -> year_name
#printing result
print(paste("They began naming hurricanes and other storms in year ",year_name))
```






# Question 2: Mapping

#### a.	plot all the storms in the database atop a simple world map.

```{r}
#install.packages("maps") #Installing "maps" Package

#defining a function to plot world map using 'ggplot2' and 'map' libraries
hurricaneBasemap = function() {
  
  #converting 'world' map from 'maps' package into dataframe 
  world = map_data("world")  
  
  basemap = ggplot() +
    #from 'hurrs' dataset, 
    #min(longitude)=-110,max(longitude)= 63
    #min(latitude)=7.2, max(latitude)= 81
    # So, fixing coordinates considering some offset value
    coord_fixed(xlim=c(-130,30),ylim=c(0,90)) + 
    
    #plotting world map using data from 'world' dataframe
    geom_polygon(data=world, aes(x=long,y=lat,group=group), color="gray50", fill="white")
  return(basemap) 
}
hurricaneBasemap() +
  
  #connecting observations from each hurricane id
  geom_path(data=hurrs, aes(x=longitude,y=latitude,group=id,color=windspeed)) +
  
  #Giving sequential coloring to observations as per wind speed. sequence followed: Yellow to Red
  scale_color_distiller(type="seq",direction=1,palette="YlOrRd")
```





#### b. write a function that takes a year as input and makes a similar plot for only the hurricanes during that year. The function doesn’t need to return a value, but it does need to generate a plot. Test your function by making the plot for a year of your choice.

function to plot hurricanes during specific year

```{r}
hurricane_in_year<- function(in_year){ 
  
  #filtering hurrs data as per year provided in argument
  hurrs_year <- hurrs %>% filter(year==in_year)
  
  #Plotting the hurricane data using function and plots used in question 2a
  hurricaneBasemap() +
    geom_path(data=hurrs_year, aes(x=longitude,y=latitude,group=id,color=windspeed)) +
    scale_color_distiller(type="seq",direction=1,palette="YlOrRd") +
    labs(title = paste("Hurricanes in year",in_year),x="Longitude",y="Latitude")
}
```

plotting hurricanes in year 1990

```{r}
hurricane_in_year(1990)
```






# Question 3: The local perspective

## a. Using the formula given in project introduction, write a function that calculates distance between (lon1,lat1) and (lon2,lat2). Test it.

loading word.cities database containing coordinates of cities from
'maps'package

```{r}
cities = world.cities 
```

defining a function to calculate distance from latitude and longitude
coordinates

```{r}

distance_between <- function(lon1, lat1, lon2, lat2) { 
  dx = (lon2 - lon1) * cos(pi * (lat1 + lat2) / 360)
  dy = lat2 - lat1
  
  D = 111.325 * sqrt(dx ^ 2 + dy ^ 2)
  return(D)
}
```

Distance between Leeds and Cardiff

```{r}
distance_between(-1.55, 53.81, -3.18, 51.48) 
```

*Distance between Leeds and Cardiff is*
`r distance_between(-1.55, 53.81, -3.18, 51.48)` km







#### b.	Adapt your function so that instead of (lon2,lat2) it takes the name of a city as input, finds its latitude– longitude coordinates, and returns the distance to (lat1,lon1). Test it. Give this function a new name and have it call the function from part (a).

```{r}
#defining function to calculate distance of city from coordinates 
city_distance <- function(lon1, lat1, city_name) {
  lon2 <-cities %>% filter(name == city_name) %>% select(long) %>% as.numeric()
  lat2 <-cities %>% filter(name == city_name) %>% select(lat)  %>% as.numeric()
  distance_between(lon1, lat1, lon2, lat2)
}
```

Calling the function

```{r}
#Distance between Leeds and Lisburn
city_distance(-1.55, 53.81, "Lisburn")
```

Distance between Leeds and Lisburn
`r city_distance(-1.55, 53.81, "Lisburn")` km





#### c. A common mistake in working with latitudes and longitudes is to swap them around. It isn’t always possible to catch user errors of this sort, but one test is that longitude can take values from -180 to 180 while latitude is always -90 to 90. Modify your function to check whether it’s likely that the user has swapped lat1 and lon1 and if so, provide a useful warning message.

defining function to check if entered coordinates are in correct range

```{r}
coord_check <- function(lon1, lat1, city_name) {
  if (-180 > lon1 | lon1 > 180){
    warning("Enter correct Longitude!")
  }
  else if(-90 > lat1 | lat1 > 90) {
    warning("Enter correct Latitude!")
  }
  else {
    lon2 <- cities %>% filter(name == city_name) %>% select(long) %>% as.numeric()
    lat2 <- cities %>% filter(name == city_name) %>% select(lat)  %>% as.numeric()
    distance_between(lon1, lat1, lon2, lat2)
  }
}
```

calling the function

```{r}
#checking with wrong latitude
#It should show warning:"Enter correct Latitude!"
coord_check( -3.18, 92, "Lisburn")
```





#### d.	Choose an eastern North American or Caribbean city, and use your function to determine which storms passed within 100 km of it. Plot those storms on a map like those in Q1, with a dot on top showing where the city is.

function to find storms passed within 100km radius of a city

```{r}
near_city_100km<- function(city_name){
  city_coord<- filter(cities, name == city_name) #filtering city name
  #finding city coordinates
  lon1<- city_coord$long 
  lat1<- city_coord$lat 
  #adding a distance column containing distance of city from storm
  distance_100<- hurrs %>% mutate(distance = distance_between(lon1,lat1,longitude,latitude)) %>% 
    filter(distance<100) #filtering storms within 100km radius
  return( distance_100) # returning same filtered table 
}
```

calling the function with city name New York

```{r}
near_city_100km("New York")
```

Plotting on map hurricanes within 100km radius of New York

```{r}
#function to return basemap to plot hurricanes within 100 km of city
hurricaneBasemap1 = function() {
  library(maps) #calling 'maps' library to draw geographical maps
  
  #converting 'world' map from 'maps' package into dataframe 
  world = map_data("world")  
  
  basemap = ggplot() +
    coord_fixed(xlim=c(-80,-70),ylim=c(35,50)) + 
    geom_polygon(data=world, aes(x=long,y=lat,group=group), color="gray50", fill="white")
  return(basemap) 
}

# Plotting on map hurricanes within 100km radius of New York 
hurricaneBasemap1() +
  
  #connecting observations from each hurricane id
  geom_path(data= near_city_100km("New York"), aes(x=longitude,y=latitude,group=id,color=windspeed),size=1) + 
  
  #Giving sequential coloring to observations as per wind speed. sequence followed: Yellow to Red
  scale_color_distiller(type="seq",direction=1,palette="YlOrRd") + labs(title = "Storms within 100km radius of New York",x="Longitude",y="Latitude") +
  # Green dot showing location of city  
  geom_point(data= filter(cities,name=="New York"),aes(x=long,y=lat),color="green",size=2)
```






# Question 4: Climate change

#### a.	Globally, the number of hurricane-strength storms has been clearly increasing because of climate change. Is this true in the Atlantic specifically? Plot the number of storms in each year.

Plotting all Hurricane-strength storms count over the years

```{r}
#counting total storms, hurricanes and storms which are not hurricanes from dataset.
all_storms<-hurrs %>% group_by(year) %>% summarise("all_storms"=n())
hurrs_true<-hurrs %>% group_by(year) %>% filter(isHurricane==TRUE) %>% summarise(hurricane=n())
hurrs_false<-hurrs %>% group_by(year) %>% filter(isHurricane==FALSE) %>% summarise(not_hurricane=n())

#Joing all counts in single dataset 'hurrs all'
hurrs_all <-left_join(hurrs_true,hurrs_false) 
hurrs_all <- hurrs_all %>% left_join(all_storms) 

#using gather function to combine all storm count columns to single column
df<-hurrs_all %>% gather(key = "storm_type", value = "storm_count",-year)

#Plotting all Hurricane-strength storms count over the years
ggplot() + 
  geom_smooth(data=hurrs_true,aes(x=year,y=hurricane), se=F,color="red",method = 'loess') +
  geom_line(data=hurrs_true,aes(x=year,y=hurricane),color="black") +
  labs(title = "Hurricane-strength storms count over the years", y="Hurricane Count")

```

***The plot shows that, the number of hurricane strength storms is increasing***






plotting count of Hurricane, not Hurricane and All storms counts

```{r}
ggplot(df, aes(x = year, y = storm_count)) + 
  geom_smooth(aes(color = storm_type, linetype = storm_type,),method = 'loess',se=FALSE) + 
  scale_color_manual(values = c("darkred", "steelblue","black")) +
  labs(title = "All types storms over the years", y="Storm Count")
```

***From the plot it is observed that, total number of storms is
continuously increasing over the years along with non-hurricane storms.
Relatively, There is no significant increase in Hurricane-strength
storms***





#### b.	The total number isn’t the only aspect of hurricane statistics that might change with climate change. What about the geographical area affected? Make a plot that lets you visually evaluate whether the latitude range where storms cross the coastline (make landfall) is expanding in recent decades. Does your plot suggest any concerns about data quality and data coverage in this database?

Defining a function to plot hurricane landfall from 1851 till specific year

```{r}
#creating basemap to plot on world map
hurricaneBasemap1 = function() {
  library(maps) 
  world = map_data("world")  
  basemap = ggplot() + coord_fixed(xlim=c(-120,-20),ylim=c(0,60)) +
    geom_polygon(data=world, aes(x=long,y=lat,group=group), color="gray50", fill="white")
  return(basemap) 
}
# Defining a function to plot hurricane landfall from 1851 till specific year
landfall_till<- function(till_year){
  hurrs %>% filter(isLandfall==TRUE,year<=till_year) -> df2
  plot<-hurricaneBasemap1() +
    geom_point(data = df2, aes( x = longitude , y = latitude , group = id , color = year )) +
    scale_color_distiller( type = "seq" , direction = 1 , palette = "YlOrRd" ) + 
    labs( title = paste("Hurricane landfall from 1851 to ",till_year))
  return(plot)
}
```

calling the function for making four different plots of hurrican
landfalls from 1851 till year 1920,1950,1990,2020 on world map

```{r}
library(ggpubr) # calling necessary library to combine plots
ggarrange(landfall_till(1920),landfall_till(1950), landfall_till(1990), landfall_till(2020),
          labels = c("A", "B", "C","D"),
          ncol = 2, nrow = 2)
```

**It is seen in the graph that over the years,storms making landfall are
expanding across eastern coastal regions of USA towards Mexico. Number of landfalls in Caribbean counties is also significantly increased over the years.**

**Data Quality: It looks quiet strange that hardly any hurricane hit any Caribbean country or Mexico from year 1851 to 1920. This raises a question whether data of hurricane landfalls before 1920 is included in this data set or not.**






#### c.	Is peak wind speed increasing over time? Answer this using whatever kind of plot or table you think is most helpful.

Plotting Maximum wind speed over the years

```{r}
hurrs %>% group_by(year) %>% 
  summarise(max.wind=max(windspeed,na.rm = TRUE)) %>% 
  ggplot() +
geom_smooth(aes(x=year,y=max.wind,color=year),color='blue',se=F) + geom_line(aes(x=year,y=max.wind),color='red')+
  labs(title = "Maximum wind speed over the years", x="years", y="max wind speed")
```

***The plot shows that maximum wind speed of Hurricanes each year is increasing over the years***





#### d.	What about the seasonal timing of hurricanes? First, make a histogram of storms by month over the entire dataset. This should show that there is a distinct hurricane season—they tend to happen at a certain time of year. Use a bin width of 1 month.

Seasonal Timings of Hurricanes

```{r warning=FALSE}
hurrs %>% 
  group_by(month) %>% 
  summarise(count = n()) %>% # Calculate count for each month
  ggplot(aes(x = month, y = count, fill = factor(month))) + 
  geom_bar(stat = "identity", color = "white") +
  scale_x_discrete(limits = 1:12) +
  theme_minimal() +
  labs(title = "No. of storms in each Month")
```

***This plot shows that 8th, 9th,10th month of the year is seasonal
timing of Hurricanes***






#### e.	add a “decade” column to the hurrs dataframe.

adding "decade" column to the hurrs dataframe

```{r}
hurrs<- mutate(hurrs,decade=year-year%%10)
```






#### f. make a plot that shows whether “hurricane season” is expanding into earlier and later months of the year over time. 

Is "hurricane season" is expanding into earlier and later months of the
year over time?

```{r warning=FALSE}
hurrs %>% group_by(month) %>% ggplot() + geom_count(aes(x=year,y=month)) + labs(title = "Hurricane occurances in different months of each year")+ scale_y_discrete(limits=1:12)
```

***The plot indicates that the hurricane season which usually observed in 8th,9th and 10th month of the year is over the time expanding into 7th , 6th month and 5th month (earlier months of the year)***
